---
title: "Getting and Cleaning Data Course - data table package"
author: "Lucy Njoki"
date: "3/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls())

library("data.table")
```

Creating a data frame
```{r}
DF = data.frame(x = rnorm(9), y = rep(c("a", "b", "c"), each = 3),
                z = rnorm(9))
head(DF)
```

Creating a data table
```{r}
DT = data.table(x = rnorm(9), y = rep(c("a", "b", "c"), each = 3),
                z = rnorm(9))

head(DT)
```

Subsettibg Rows
```{r}
DT[2,] #row 2

DT[DT$y == "a",]

DT[c(2,3),] #rows 2 and 3
```

Subsetting Columns
```{r}
DT[, c(2,3)] #columns 2 & 3
```

Calculating values for varibles with expressions
```{r}
DT[, list(mean(x), sum(z))]

DT[,table(y)]
```

Adding new columns
:=  - used to add a new varibale to the data table
```{r}
DT[, w:=z ^2]

DT1 <- copy(DT)
head(DT1,3)

DT[,m:={tmp <- (x+z); log2(tmp +5)}] #m = log2(x+z)+5

DT[,a:= x>0] #plyr  like operations

DT[,b:= mean(x+w), by  = a] #mean(x+w) when a = true, and place it all rows when a=false and vice versa
```

Special variables in data table package that allows fast applications
.N - An integer of length 1 and contains the number of times a particular group appears.
```{r}
set.seed(123)
dt <- data.table(x = sample(letters[1:3], 1E5, TRUE))
dt[, .N, by = x] #counts the number in which the letters 1 to 3 appear grouped by x
```

KEYS

1. Subsetting on the basis of a certain variable

```{r}
DT <- data.table(x = rep(c("a", "b", "c"), each = 100),
                 y = rnorm(300))
setkey(DT, x) #subsetting the key to be x

DT['a'] # subsetting values to only those equal to a
```

2. Join TWO data tables

```{r}
DT1 <-  data.table(x = c("a", "a", "b", "d+1"), y = 1:4)
DT2 <- data.table(x = c("a", "b", "d+2"), z = 5:7)

#settng x as the key
setkey(DT1, x); setkey(DT2, x)

merge(DT1, DT2)

```

Fast Reading compared to a data frame
```{r}
big_df <- data.frame(x = rnorm(1E6), y = rnorm(1E6))

file <- tempfile() #creatiing temporary file

write.table(big_df, file = file, row.names = FALSE,
            col.names = TRUE, sep = "/t", quote = FALSE)

system.time(fread(file)) #time taken to read the data table file

system.time(read.table(file, header  = T, sep = ":")) #time taken to read the data frame file
```


