---
title: "Sorting and Subsetting"
author: "Lucy Njoki"
date: "4/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls())
```


#Subsetting Data

One can create a subset of data by:

>specific value;

>variable;

>on rows and columns

Also logical statements (**And(&)** and __Or(|)__) can be used to subset data

```{r Subsetting}

set.seed(20200401)

x <- data.frame("var1" = sample(1:5), "var2" = sample(6:10),
                "var3" = sample(11:15))

#scrambling values so that they dont appear in specific order & introduding NA
x <- x[sample(1:5),]; x$var2[(1:3)] = NA
x

#subset by specific value
x[,1]

#subset by variable
x[, "var1"]

#subset on rows and columns
x[1:2, "var2"]
```

_set.seed() used to get reproducible sample_

```{r Subset using logical (Ands & Ors)}

x[(x$var1 <= 3 & x$var3 > 11),] #and

x[(x$var1 <= 3 | x$var3 > 15), ] #or
```

```{r Dealing with missing data}

x[which(x$var2 > 2),]
```
_which() does not return NA, therefore the a good way to subset data when NAs are present._

#Sorting Data

3 functions used:

>sort()

>order()

>plyr::arrange()

```{r Sorting Data}
#using sort()

sort(x$var1) #increasing

sort(x$var1, decreasing = T) #decreasing

sort(x$var2, na.last = TRUE) # NAs to be last

#using ordering()

x[order(x$var1),] # orders the DF, by var1 to be in increasing order

x[order(x$var1, x$var3),]

#using pylr::arrange()

library(plyr)

arrange(x, var1) #data is arranged by var1 increasing

arrange(x, desc(var1)) #Var1 is in decreasing
```

#Adding rows and columns in a DF

>data$newVaraible

>cbind() - adds a column to the right of the DF

>rbind() -  adds a row to the bottom of the DF

```{r Adding rows and columns in a DF}
x$var4 <- rnorm(5)

#cbind()

y <- cbind(x, var4 = rnorm(5));y

#uding rbind()

z <- rbind(x, rnorm(5));z
```